Can historical data predict population responses to climate change experiments?

### 2016-12-23 precip/manuscript/precip_ms.tex
adding discussion revision

\smallskip \textbf{Authorship statement:} ARK and PBA designed the experiment and supervised data collection. AT helped with the data analysis. JB and CA helped with the soil moisture model. 
\smallskip

@@

Climate is an important driver of population ecology however there have been few tests of whether observational data correlating population performance with climate variation can be used to predict how populations respond to climate change experiments. We tested whether longterm observational data relating the effects of climate variation to the survival, growth and recruitment of four common plants in a sagebrush steppe could be produce accurate predictions for how each species would respond to a drought and irrigation experiment.

We used rainout shelters and automatic sprinklers to manipulate the soil moisture in sixteen new plots established at the same field site as the longterm observational data. After running the experiment for five years, we analyzed how the growth, survival and recruitment of each species responded to the experimental drought and irrigation treatments. We made predictions using two types of models fit to the observational data collected prior to the experiment: a no climate model that only included the effects of plant size and competition on plant performance in each plot in each year; and a climate model that also included the effects of three seasonal soil moisture variables. We then used these models to predict the performance of each species during the course of the experiment. We also generated one step a head predictions of population size in each experimental plot using an individual based population model. We compared predictions made by the the no climate and climate models to the observed experimental responses.

Over the course of the experiment, average cover of the perennial grasses, \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly in the drought treatment. At the level of individual vital rates, experimental drought reduced the survival of \textit{Hesperostipa comata} and \textit{P. spicata}  and the growth of the grass \textit{Poa secunda}. In contrast, drought increased the growth of the shrub \textit{A. tripartita}. The climate model better predicted the response to the experiment than the no climate model in six out of twelve cases. Across all species and vital rates, there was a strong positive correlation between the observed effects of the experiment and the effects predicted by the climate model fit to the observational data. At the population-level, the climate model made better predictions of the changes in species cover than the no climate model for \textit{P. secunda} and \textit{P. spicata}.

Observational climate data held valuable information for predicting species' responses to a climate change experiment in this ecosystem. Treatment responses often matched the direction of predicted responses even when the effects were not significant. We were better able to predict species' responses to the drought treatment than to the control and irrigation treatments. This suggests that soil moisture is an important factor in predicting the population dynamics of these species but only when water is truly limiting.

@@

Ecologists often resort to one of two methods for predicting the effects of future climate change on populations and communities: they may use experiments to manipulate aspects of climate directly and observe the response of populations \citep{elmendorf_experiment_2015,knapp_pushing_2016,compagnoni_warming_2014}; or they may use longterm observational data on species performance and abundance collected over many years and relate this to ambient annual variation in climate  \citep{koons_climate_2012,lunn_demography_2016,dalgleish_climate_2010,jenouvrier_demographic_2009}. The strength of the experimental approach is in the stronger inference that comes from manipulating some aspects of climate while controlling for other factors; for instance knowing that loss of snow cover, and not necessarily changes in soil temperature or moisture are the factors causing a species performance to change with warming \citep{compagnoni_warming_2014}. It also allows for the creation of conditions that may be more extreme than those observed historically \cite{knapp_pushing_2016} but are possible in the future. However, it is often expensive to control climate at even the smallest scales, and larger scale climate manipulation is often impossible.  Moreover, experimental manipulation can come with artifacts that may make them less then ideal models for understanding and predicting the effects of future variation in climate \citep{wolkovich_warming_2012}.

Using observational data to predict the effects of climate variation on populations has many advantages over using experiments. The primary one being the ever increasing availability of long-term ecological data and ever increasingly detailed climate data. Observational studies may also be the only way to study the effects of climate on large and or migratory species, for which it would be difficult to manipulate climate \citep{koons_climate_2012,jenouvrier_demographic_2009,aubry_climate_2013}. However, there are many potential drawbacks to using essentially correlative relationships between species performance and climate to predict future species' responses. First, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}. \citep{teller_linking_2016} estimate that even cutting edge statistical approaches for fitting relationships between climate and species performance require at least 20-25 years of independent climate observations before they perform well. Moreover, even strong correlations between species performance and the climate covariates we choose to include in our models may not be due to direct causation and may not hold up in future years of observation \citep{hilborn_correlation_2016}.

And yet even overcoming this challenge, predicting the effects of climate change would face yet another challenge: in many parts of the world, climate change will lead to mean annual temperatures that will be above the range of historical variation. If species performance responds non-linearly to temperature, we may be unable to predict how species respond to temperatures outside the range of observed variation \citep{doak_demographic_2010}. In addition, climate change will not just change the average climate, but is also likely to increase the variance in precipitation and lead to extreme events, which will have their own consequences independent of changes in climate means \citep{gherardi_enhanced_2015,jentsch_new_2007}. Any models based on observations drawn from the historical range of variation will therefore be extrapolating beyond both the range of observed averages and variance when used to predict the future \citep{williams_novel_2007}. Demonstrating that models fit to historical data can accurately predict the effects of experimental climate manipulations, especially manipulations that generate extreme conditions, would be strong confirmation that the climate effects they describe are not spurious and will hold even in the novel conditions of the future \cite{adler_can_2013}.

There is abundant evidence that plant performance shows high year to year variation, both at the level of individual growth, survival and reproduction and total population abundance. Precipitation is often especially important in driving interannual variation in net primary productivity among terrestrial plants \citep{knapp_variation_2001,hsu_anticipating_2014}, the annual growth rings in trees \citep{yang_3500-year_2014} and smaller plants \citep{srur_annual_2009,franklin_growth_2013}, as well as the germination and reproductive output of many annuals \citep{venable_bet_2007}.

Despite the clear signs that precipitation drives net primary productivity at the ecosystem level and annual variation in individual growth rate in many plants, there have been relatively few studies that clearly link observed climate variation to species performance in population models \citep{ehrlen_advancing_2016}. Fewer still have tested whether the population-level effects of climate variation shown in observational data can actually be used to predict the responses of plant populations to future climate variation. \citep{adler_can_2013} showed that population models based on observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland. Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.  However, another three species, showed responses to the experimental conditions that were not well predicted by historical observations. The authors suggested that limited replication in the historical data for two of these species and changing competitive conditions in the community may have led to the poor predictions.

The demography of and competitive interactions between three dominant perennial bunchgrasses and a shrub in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least seven different studies since 2010 \citep{adler_coexistence_2010,adler_forecasting_2012,adler_weak_2009,adler_weak_2016,chu_direct_2016,chu_large_2015,dalgleish_climate_2010}. And several of these studies report significant effects of climate variation on the vital rates and overall population growth rates of these species \citep{dalgleish_climate_2010,adler_forecasting_2012,adler_weak_2009,chu_direct_2016}. This well-studied system offers the ideal opportunity to test whether statistical associations between annual climate and plant demography in longterm observational data can be used to predict the responses of plant populations to experimental climate change experiments.

In this study, we report how the four dominant plant species at the USSES respond to a five year drought and irrigation experiment and use the results to address two research questions: first, how much do the growth, recruitment and survival of our target species differ between the precipitation manipulation treatments? If our experiment does affect species vital rates we interpret that as strong evidence that changes in precipitation should have an effect on populations in the future. Second, can we predict each species' response to the experimental conditions based on how they respond to natural climate variation in the observational data? If models based on observational data can predict the response of species to this experiment this should give us more confidence in longterm population monitoring data to predict species response to climate change.

@@

Our analysis consists of two separate datasets and three different categories of vital rate models. We refer to the first dataset as the observational data. It consists of all the historical data collected from 1925 to 1957 as well as the modern data collected from the same plots from 2007 to 2010. These data record the response of plants in each plot to the ambient climate variation. We refer to the second dataset as the experimental data.  It consists of the data collected from 2011 to 2016 from the 16 new experimentally manipulated plots, as well as from 14 of the original historical plots which serve as ambient climate controls.

In order describe the effects of the experimental treatments on each vital rate, we fit "treatment" models. The treatment models included parameters for the effects of the drought and irrigation treatments on each vital rate. We fit these models to both datasets together, all the experimental and all the observational data. We did this because we wanted to focus our predictions on the effects of the experimental treatments on the vital rates, rather than any differences between the datasets in the effects of crowding and plant size on the vital rates.

Next in order to test how well we could predict the responses in the experimental plots, we fit two classes of models to the observational dataset only. Thus when we used these models to predict the response of each species to the five years of experimental data we are generating predictions for out of sample data. We refer to the first type of model as the "no climate" models. These models include parameters for the effects of competition on each vital rate and the size dependence of survival and growth but they do not include climate or treatment effects. Lastly, we fit the "climate" models which include the effects of annual variation in soil moisture on each vital rate.  The no climate model provides us a baseline by which to measure the accuracy of the predictions from the climate model. Because much of the variation in growth, survival and recruitment in this system can be explained by plant size and competition, we expect that these two models will make similar predictions for individual plant performance in the experiment.  However, if the climate model makes more accurate predictions than the no climate model, this indicates that the climate parameters it includes contain useful information for prediction.

@@

Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species responses to experimental climate manipulations (fig \ref{fig:parPredictions}). This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models (fig \ref{table:overallPreds}). This should give us some hope that even when climate effects in demographic models fit to observational data are weak or not significant, they may contain useful qualitative information on the direction of climate effects in the future.

Although a number of previous studies of the plants in this system have looked the effects of climate on demography and survival, this is the first study to demonstrate effects of climate experimentally. The significant decline in \textit{H. comata} cover we report in the experiment (fig \ref{fig:coverChange}) is consistent with the population-level sensitivities to precipitation estimated by previous studies (\citep{chu_direct_2016},\citep{dalgleish_climate_2010}, \citep{adler_forecasting_2012}). Each of these studies report \textit{H. comata} having the strongest positive population response to increased precipitation among all the species; this is matched by our finding that this species has the strongest negative response to experimental drought (fig \ref{fig:growthTreat}, \ref{fig:survivalTreat}). On the other hand if we had only conducted an irrigation experiment our results may not have shown this consistency with previous work as \textit{H. comata} showed no positive response to irrigation. Previous work has also generally reported postive population responses to precipitation for the other grasses, \textit{P. secunda} and \textit{P. spicata} (\citep{adler_forecasting_2012}, \citep{chu_direct_2016}, \citep{dalgleish_climate_2010}). These results are consistent with significant decrease we saw in \textit{P. spicata} cover and survival and \textit{P. secunda} growth and survival in the drought plots in our experiment (figs \ref{fig:coverChange}, \ref{fig:growthTreat}, \ref{fig:survivalTreat}). The magnitudes of the negative drought effects were greater than the magnitudes of the positive irrigation effects on these grasses.

In contrast to the effects on the grasses, the effects on \textit{A. tripartita} are more complicated. The effects of drought and irrigation on \textit{A. tripartita} growth were highly dependent on plant size (figs \ref{fig:growthTreat}): drought had a positive effect except possibly for the largest plants (fig \ref{fig:parPredARTRGrowth}). This result matches the negative effects of precipitation on this species estimated by \citep{adler_forecasting_2012} and \citep{chu_direct_2016}. These studies also report relative strong indirect effects of precipitation on \textit{A. tripartita} mediated by its competition with grasses. It is possible that some of the positive effect on \textit{A. tripartita} growth we see in the drought plots is the result of reduced grass competition these plots (fig \ref{fig:coverPred}; \citep{chu_direct_2016}). However, our growth model includes interspecific crowding and so should take into account any changes in grass abundance that could be driving a positive response from \textit{A. tripartita}. So this leaves us with the question of why this species would show a positive direct response to drought? Although there is some evidence that saturated soils can be detrimental to big sagebrush (\textit{A. tridentata}), a closely related species \citep{sturges_response_1989}, there was no evidence in our data that soils were saturated with moisture in the ambient or irrigation treatments (fig \ref{fig:dailyVWC}). In a similar study to ours, \citep{germino_desert_2014} showed that big sagebrush increased in response to deep soil moisture but actually decreased in response to shallow soil moisture. Perhaps \textit{A. tripartita} responds similarly and benefits directly from the drier soils. Another possible explanation is that our drought treatments reduced snow cover in the winter and early spring, an effect that has been shown to benefit big sagebrush in other ecosystems \citep{perfors_enhanced_2003}.

One trend that appears to stand out in these results is the relatively asymmetrical effects of drought and irrigation. Although our irrigation plots raised soil moisture, we saw little positive effect of added moisture on the same species that were reduced by drought (figs \ref{fig:growthTreat}, \ref{fig:survivalTreat}). We expected that in this water limited system, increasing soil moisture would lead to a stronger increase in plant performance. But perhaps we should not have been surprised by this result. It is possible that adding water to this system does little to increase grass growth and survival for a number of reasons.  First, these species are presumably able to escape drought, by growing early in the year, or avoid drought stress later in the year through high water use efficiency \citep{bazzaz_physiological_1979,franks_plasticity_2011}. If the species in our system are well adapted to the ambient water limitation, it makes some sense that they would show relatively small responses to increased moisture. Indeed our soil moisture data generally show a pulse of soil moisture in the spring when many grasses are actively growing (fig \ref{fig:dailyVWC}). During this window of active growth soil moisture in sagebrush steppe may be quite high and perhaps other resources become limiting \citep{caldwell_rapid_1987}. Moreover, our treatments had relatively smaller effects on soil moisture during the spring (table ).

A similar result was recently reported from another precipitation experiment in a warm desert grassland: the effect of increased soil moisture on grass productivity in that system showed a concave down pattern of response to increased soil moisture \citep{gherardi_enhanced_2015}.  Increasing precipitation had a weaker effect on annual grass productivity than decreasing precipitation by an equal amount. A pattern like this is qualitatively similar to our finding significant decreases in either the cover or performance of the three grasses in drought plots, but not seeing any statistically significant increase in performance in the irrigated plots.

Reflecting on the possible mechanisms by which soil moisture could have affected these species may also shed some light on our sucesses and failures at prediction. The drought effects we observed on the three grasses were often stronger than the effects we predicted, while the irrigation responses observed were often weaker than predicted (figs \ref{fig:parPredHECOGrowth}, \ref{fig:parPredHECOSurvival}, \ref{fig:parPredPOSEGrowth}, \ref{fig:parPredPOSESurvival}, \ref{fig:parPredPSSPGrowth}, \ref{fig:parPredPSSPSurvival}). If we had fit our growth and survival models with a non-linear function for soil moisture, perhaps informed by more mechanistic understanding of water limitation in this system, we may have made more accurate predictions of the drought and irrigation effects \citep{ehrlen_advancing_2016}.

Out of all the climate effects we predicted and observed, we only observed four cases where vital rate predictions and observations were both significantly different from zero (fig  \ref{fig:parPredictions}). In three of these cases, we successfully predicted the direction of the treatment effects. However, for \textit{P. secunda} recruitment we predicted a response to irrigation that was significantly positive and yet observed a significantly negative response (figs \ref{fig:parPredictions}). This is arguably our greatest error in prediction. But there may be some mitigating factors that can excuse our error. First we note that we observed decreases in recruitment in both the drought and irrigation plots for \textit{P. secunda} and also for \textit{P. spicata} (fig \ref{fig:recruitmentTreat}). So while we observed a significant decrease in \textit{P. secunda} recruitment in the irrigated plots this may have been due to underlying differences in the experimental plots from the historical control plots. Because recruitment is modeled at the quadrat level we also had fewer observations and predictions to work with for these recruitment models than we had for the other vital rates.

Using the IBM, we demonstrated that the climate model predictions of the vital rates could be used to generate population-level predictions that were also better than the no climate models in some cases: the climate model improved one step ahead predictions of cover changes of \textit{P. spicata} and \textit{P. secunda} (table \ref{table:corPGR}). In the drought treatment our one step ahead cover predictions for \textit{H. comata} and \textit{P. secunda} were also better than the no climate model. In this analysis, we either used the climate model, or the no climate model to generate predictions for the IBM. In theory we could have refined our predictions by  climate model to predict all three vital rates in the IBM. Improvements of our cover predictions may have been possible if we only included climate models in the IBM when those models improved vital rate predictions--for instance, the \textit{A tripartita} IBM would only include the climate model for growth but not for survival and recruitment.

Mixed success at predicting species responses to high and low moisture availability has interesting implications.  On the one hand it is reassuring that observational data captures important climate effects. On the other hand, will knowing that some species' responses are unpredictable, but not knowing which ones, discourage ecologists from making any predictions at all? Among plants, perhaps improved ecophysiological models or trait data can be used as a screen to generate hypothesis about which species are more likely to be affected by climate and in which ways. In our system, for instance, it may make sense that the grasses showed a stronger response to the drought treatment than the a woody shrub. Although these grasses are adapted to the arid conditions that characterize the sagebrush steppe, they thrive during the brief window in spring and early summer when the soil moisture and temperatures are warmer.  \textit{A. tripartita} on the other hand grows throughout the summer and has deeper roots than the grasses, these traits may buffer it from some of the annual variability in soil moisture experienced by the grasses.

Will our predictions be useful for the predicting the effects of climate change in the future? Because these species compete, one could argue that our predictions for any one species in this community will only be as good as the predictions we make for their competitors, pathogens and predators \citep{tylianakis_global_2008}. For instance, while we observed weak treatment effects on \textit{A. tripartita} in our experiment (fig \ref{fig:coverChange}) could drought ultimately lead to a decline in the cover of its competitors which will cause it to increase in abundance in the future?  While this is possible for some species, research at our site and several other grasslands suggest that species are more limited by intra-specific competition than inter-specific competition \cite{chu_direct_2016,adler_weak_2016}. In theory, this should mean that we can predict species short term responses to climate change with single species models (\citep{levine_competitors_2010,kleinhesselink_indirect_2015,adler_forecasting_2012}).

Our results give us more confidence that historical observational data can be used to detect and predict the demographic effects of climate change among sagebrush steppe plants. This should encourage more researchers to try and use observational data to predict the future in both experimental and natural settings \citep{houlahan_priority_2016}. Nevertheless, our success at predicting the short-term response of two species to a small-scale climate manipulation is not likely to be very reassuring to applied ecologists and resource managers wishing to make accurate predictions about the effects of climate change in large complex systems. Clearly more work is needed to learn how best to meet the challenge of prediction in ecological systems. Towards that goal, perhaps the best way forward is to conduct more tests like this one.

>>>>

\smallskip \textbf{Authorship statement:} ARK and PBA designed the experiment and supervised data collection. AT helped with the data analysis. JB and CA helped with the soil moisture model. ARK wrote the first draft of the manuscript which all authors edited.
\smallskip

@@

Climate is an important driver of population dynamics and annual variation in demographic rates often correlate with variation in weather. However, the predictive potential of such correlations is largely unknown. We tested how well population models fit using long-term observational could predict the responses of four dominant plant species in a sagebrush steppe to a drought and irrigation experiment.

We used rainout shelters and automatic sprinklers to manipulate the soil moisture in sixteen plots established at the same field site where long-term observational data was collected. We report how the growth, survival and recruitment of each species responded to the experimental drought and irrigation treatments after five years. We then attempted to predict these treatment responses using two models fit to the observational data collected prior to the experiment: a no climate model that only included the effects of plant size and competition on plant performance in each plot in each year, and a climate model that also included the effects of three seasonal soil moisture variables. We compared predictions made by the no climate and climate models to the actual experimental responses. We also generated one-step-ahead predictions of population size in each experimental plot using individual based population parameterized with the vital rate estimates from the climate and no climate models.

Over the course of the experiment, average cover of the perennial grasses, \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly in the drought treatment. At the level of individual vital rates, experimental drought reduced the survival of \textit{Hesperostipa comata} and \textit{P. spicata} and the growth of the grass \textit{Poa secunda}. In contrast, drought increased the growth of the shrub \textit{A. tripartita}. The climate model made better predictions of the experimental responses than the no climate model in six out of twelve cases. Across all species and vital rates, there was a strong positive correlation between the observed responses to the treatments and the responses predicted by the climate model. At the population-level, the climate model predicted changes in species cover more accurately than the no climate model for \textit{P. secunda} and \textit{P. spicata}.

Observational climate data held valuable information for predicting species' responses to a climate change experiment in this ecosystem. Treatment responses often matched the direction of predicted responses even when the effects were not significant. We were better able to predict species' responses to the drought treatment than to the control and irrigation treatments, suggesting that soil moisture is an important factor for predicting the population dynamics of these species but only when water is truly limiting.

@@

Ecologists often resort to one of two methods for predicting the effects of future climate change on populations and communities: they may use experiments to manipulate aspects of climate directly and observe the response of populations \citep{elmendorf_experiment_2015,knapp_pushing_2016,compagnoni_warming_2014}; or they may use long term observational data on species performance and abundance collected over many years and relate this to ambient annual variation in climate  \citep{koons_climate_2012,lunn_demography_2016,dalgleish_climate_2010,jenouvrier_demographic_2009}. The strength of the experimental approach is in the stronger inference that comes from manipulating some aspects of climate while controlling for other factors; for instance, knowing that loss of snow cover, and not necessarily changes in soil temperature or moisture are factors causing a species performance to change with warming \citep{compagnoni_warming_2014}. It also allows for the creation of conditions that may be more extreme than those observed historically \cite{knapp_pushing_2016} but are possible in the future. However, it is often expensive to control climate at even the smallest scales, and larger scale climate manipulation is often impossible.  Moreover, experimental manipulation can come with artifacts that may make them less then ideal models for understanding and predicting the effects of future variation in climate \citep{wolkovich_warming_2012}.

The use of observational data to predict the effects of climate variation on populations has its own advantages and disadvantages. One important advantage is cost: analyses of already existing long-term ecological data and ever increasingly detailed climate data are cheaper than experiments. Observational studies may also be the only way to study the effects of climate on large and or migratory species, for which it would be difficult to manipulate climate \citep{koons_climate_2012,jenouvrier_demographic_2009,aubry_climate_2013}. The principal disadvantage is the reliance on essentially correlative relationships between species performance and climate to predict future species' responses. First, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}. \citep{teller_linking_2016} estimate that even cutting edge statistical approaches for fitting relationships between climate and species performance require at least 20-25 years of independent climate observations before they perform well. Moreover, even strong correlations between species performance and the climate covariates we choose to include in our models may not reflect direct causation, leading to failures when predicting future, out of sample performance \citep{hilborn_correlation_2016}.

The extrapolation of climate-demography correlations presents another potential problem. In many systems, future precipitation and temperature will fall outside the range of historical variation. If species performance responds non-linearly to these drivers, fitting linear models for species responses to climate may produce larger errors when future conditions are outside the range of observed variation \citep{doak_demographic_2010}. In addition, climate change will not only alter average weather, but is also likely to increase the variance in precipitation and the frequency of extreme events, which will have their own consequences independent of changes in means \citep{gherardi_enhanced_2015,jentsch_new_2007}. Any models based on observations drawn from the historical range of variation will therefore be extrapolating beyond both the range of observed averages and variances when used to predict the future \citep{williams_novel_2007}.

Here, we combine the strengths of experimental and observational approaches by testing the ability of models fit to historical data to predict the effects of experimental climate manipulations that generate extreme conditions. A demonstration that the observational approach can skillfully predict experimental responses would provide strong confirmation that observed climate-demography correlations are not spurious and will hold even in novel conditions in the future \cite{adler_can_2013}. \citep{adler_can_2013} showed that population models based on observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation. Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.  However, the responses of another three species, were not well predicted by historical observations.

Among plant populations, interannual variation in precipitation and or soil moisture often drives variation in net primary productivity \citep{knapp_variation_2001,hsu_anticipating_2014}, the annual growth rates of the woody tissue in trees and shrubs (\citep{yang_3500-year_2014},\citep{srur_annual_2009,franklin_growth_2013}), and the germination and reproductive output of annuals \citep{venable_bet_2007}. Despite clear signs that precipitation shold be important for plant populations, there have been relatively few studies that clearly link observed variation in precipitation to species performance in population models \citep{ehrlen_advancing_2016}.

The sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho offers an ideal opportunity to test whether the climate effects in plant populations models derived from observational data can also be used to predict species responses to controlled precipitation experiments. The demography of three perennial bunchgrasses and a shrub species at the USSES have been described in detail in seven different studies since 2010, several of which report significant effects of seasonal precipitation on the vital rates and overall population growth rates of these species (\citep{adler_coexistence_2010,adler_forecasting_2012,adler_weak_2009,adler_weak_2016,chu_direct_2016,chu_large_2015,dalgleish_climate_2010}).

In this study, we report how the four dominant plant species at the USSES respond to a five year drought and irrigation experiment and use the results to address two research questions: first, how much do the growth, recruitment and survival of our target species differ between the precipitation manipulation treatments? Significant experimental effects on species vital rates imply that future changes in precipitation will impact populations. Second, can we predict each species' response to the experimental conditions based on how they respond to natural climate variation in the observational data? If models based on observational data can predict the response of species to this experiment we will gain confidence in using long-term population monitoring data to predict species responses to future climate change.

@@

Our analysis consists of two separate datasets and three different categories of vital rate models. We refer to the first dataset as the observational data. It consists of all the historical data collected from 1925 to 1957 as well as the contemporary data collected from the same plots from 2007 to 2010. These data record the response of plants in each plot to the ambient climate variation. We refer to the second dataset as the experimental data.  It consists of the data collected from 2011 to 2016 from the 16 new experimentally manipulated plots, as well as from 14 of the original historical plots which serve as ambient climate controls.

In order describe the effects of the experimental treatments on each vital rate, we fit "treatment" models. The treatment models included parameters representing the effects of the drought and irrigation treatments on each vital rate. We fit these models to using all the experimental and all the observational data. We combined the datasets because we wanted to focus our predictions on the effects of the experimental treatments on the vital rates, rather than any differences between the historical and contemporary periods in the effects of crowding and plant size on the vital rates.

Next, in order to test how well we could predict the responses in the experimental plots, we fit two classes of models to the observational dataset only.  The "no climate" models include parameters for the effects of competition on each vital rate and the size dependence of survival and growth but they do not include climate or treatment effects. The "climate" models include the effects of annual variation in soil moisture on each vital rate. The no climate model provides us a baseline by which to measure the accuracy of predictions from the climate model. Because much of the variation in growth, survival and recruitment in this system can be explained by plant size and competition, we expect that these two models will make similar predictions for individual plant performance in the experiment.  However, if the climate model makes more accurate predictions than the no climate model, this indicates that the climate parameters it includes contain useful information for prediction. Note that because these models are fit using only the observation dataset, when we use  these models to predict experimental responses we are generating true out-of-sample predictions.

@@

Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species responses to experimental climate manipulations (fig \ref{fig:parPredictions}). The historical climate-demography correlations helped predict the direction of experimental responses even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models (fig \ref{table:overallPreds}). This should give us some hope that even when climate effects in demographic models fit to observational data are weak or not significant, they may contain useful qualitative information on the direction of climate effects in the future.

\subsection*{Comparison of experimental and natural climate effects}
While previous studies in this system used the observational data to describe the effects of climate on demography and survival, this is the first study to demonstrate effects of climate experimentally. While previous studies of this system fit a variety of climate effects and used different modeling approaches, we see many points of similarity between these studies in the responses of the four dominant species to precipitation (\citep{chu_direct_2016},\citep{dalgleish_climate_2010}, \citep{adler_forecasting_2012}). First, in all three studies the strongest positive effects of precipitation among the four species are reported for \textit{H. comata}; this matches the negative effects of our drought experiment on this species (fig \ref{fig:coverChange}). This effect is driven by a negative growth and survival response to drought (fig \ref{fig:growthTreat}, \ref{fig:survivalTreat}). On the other hand, if we had only conducted an irrigation experiment our results may not have shown this consistency with previous work as \textit{H. comata} showed no positive response to irrigation. Previous studies also reported positive effects of precipitation on the other grasses, \textit{P. secunda} and \textit{P. spicata}. Again our results are consistent with this result: drought led to declines in cover of \textit{P. spicata}, \ref{fig:coverChange}, and in the growth and survival of \textit{P. secunda} (figs \ref{fig:growthTreat}, \ref{fig:survivalTreat}). As for the \textit{H. comata} the magnitudes of drought effects were greater than the irrigation effects on these grasses.

The effects of precipitation on \textit{A. tripartita} are more complicated. Previous research reported negative direct effects of precipitation on this species (\citep{adler_forecasting_2012}, \citep{chu_direct_2016}). This effect has always seemed odd because it is hard to imagine why precipitation would have a direct negative effect on a species in this dry ecosystem. But again the largely positive (but size dependent) effects of drought treatments on \textit{A. tripartita} growth should give us more confidence in the negative effects of precipitation shown in the historical data. These studies also report relative strong indirect effects of precipitation on \textit{A. tripartita} mediated by its competition with grasses. It is possible that some of the positive effect on \textit{A. tripartita} growth in our drought plots is the result of reduced grass cover (fig \ref{fig:coverPred}; \citep{chu_direct_2016}). However, our growth model includes interspecific crowding and so should take into account any changes in grass abundance that could be driving a positive response from \textit{A. tripartita}. This leaves us with the question of why this species would show a positive direct response to drought. Although there is some evidence that saturated soils in the spring are detrimental for big sagebrush (\textit{A. tridentata}), a closely related species \citep{sturges_response_1989,germino_desert_2014}, soil saturation would conservatively seem to require soils to be above 30 or 40\% volumetric water content for several weeks, something that we did not observe (fig \ref{fig:dailyVWC}). Another possible explanation is that our drought treatments reduced snow cover in the winter and early spring, an effect that has been shown to benefit related sagebrush species in other ecosystems \citep{perfors_enhanced_2003}.

Overall we were somewhat surprised by the weak effects that reducing water availability by 50\% and increasing water availability by 150\% had in this arid system. Cross-biome studies of the relationship between precipitation and ANPP generally show that arid systems are highly sensitive to water limitation \citep{huxman_convergence_2004}. We have two explanations for the seemingly weak effects of precipitation we observed on demography. First, we measure the size of the pernnial bunchgrasses in this system by their basal cover, which may not have a strong relationship with their annual production. It is likely that we would find larger effects of precipitation on these grasses if we had a more complete measurement of aboveground biomass. Moreover, much of the growth of these species may be going into roots.

Another explanation for the weak effects of precipitation are that perennial species in this cold desert ecosystems are well adapted to tolerate drought, either through escaping drought by growing early in the year, or by avoiding drought stress later in the year through high water use efficiency \citep{bazzaz_physiological_1979,franks_plasticity_2011}. Indeed our soil moisture data generally show a pulse of soil moisture in the spring when many grasses are actively growing (fig \ref{fig:dailyVWC}). Likewise, \textit{A. tripartita} is more deeply rooted than the grasses and able to continue its growth throughout the growing season by drawing from deeper soil water \citep{germino_desert_2014}. The adaptations of native perennial plants in cold deserts could make them less sensitive to water availability than species in a more mesic ecosystem.

\subsection*{Can the past predict the future?}
Our second research question was whether we could use long term observational data on species response to precipitation to predict the response of each species to the experiment. Using the IBM, we to generate predicted changes in population size for each year, we found that climate model predictions were indeed better than the no climate models for two species: \textit{P. spicata} and \textit{P. secunda} (table \ref{table:corPGR}). In the drought treatment our one step ahead cover predictions for \textit{H. comata} and \textit{P. secunda} were also better than the no climate model. Moreover, we also found that climate models produced better predictions of species vital rates for half of the species/vital rate combinations we tested (table \ref{table:overallPreds}). The rate of overall improvement in predictive ability produced by the climate models over the no climate models was similar to the results reported by \citep{adler_can_2013} who also reported improved population-level predictions for half of the species predicted. Likewise, in a within sample cross-validation analysis, \citep{tredennick_we_2016} found that including climate covariates improved population level predictions for two out of four species in a mixed grass prairie in Montana.

We also compared the treatment parameters from the treatment model fit to the experimental data to the treatment parameters predicted by the climate model fit only to the observational data \ref{fig:parPredictions}. Among all the climate effects we predicted and observed, there were only four cases where vital rate predictions and observations were both significantly different from zero (fig \ref{fig:parPredictions}) and three of these cases we successfully predicted the direction of the treatment effects. However, for \textit{P. secunda} recruitment we predicted a positive response of irrigation, but observed a negative response (figs \ref{fig:parPredictions}). From a statistical standpoint this is our arguably our greatest error in prediction. However, recruitment decreased in both the drought and irrigation plots for \textit{P. secunda} and also for \textit{P. spicata} (fig \ref{fig:recruitmentTreat}). So its likely the decrease in \textit{P. secunda} recruitment in the irrigated plots was due to underlying differences in the set of experimental plots from the historical control plots rather than the precipitation treatments.

The drought effects we observed on the three grasses were often stronger than the effects we predicted, while the irrigation effects observed were often weaker than predicted (figs \ref{fig:parPredHECOGrowth}, \ref{fig:parPredHECOSurvival}, \ref{fig:parPredPOSEGrowth}, \ref{fig:parPredPOSESurvival}, \ref{fig:parPredPSSPGrowth}, \ref{fig:parPredPSSPSurvival}). In this water limited system, we expected that experimental irrigation would lead to increases in plant performance, but we saw few cases where irrigation benefited any of the plants. A pattern qualitatively similar to this shows up in both natural and experimental data comparing precipitation to ANPP: decreases in grassland ANPP induced by drought are often of greater magnitude than increases in ANPP induced by experimental irrigation or by above average precipitation (\citep{hsu_anticipating_2014,gherardi_enhanced_2015}). If we had fit our growth and survival models with a non-linear function for soil moisture, perhaps informed by more mechanistic understanding of water limitation on the physiology of these plants, we may have made more accurate predictions of the drought and irrigation effects \citep{ehrlen_advancing_2016}.

\subsection*{Conclusion}

Our results give us more confidence that observational data can be used to detect and predict the effects of annual soil moisture variation on sagebrush steppe plants. This should encourage more researchers to try and use observational data to predict population response to climate in both experimental and natural settings \citep{houlahan_priority_2016,ehrlen_advancing_2016}. Nevertheless, our success at predicting the short-term response of two out of four species to a simple precipitation manipulation is not likely to impress applied ecologists and policymakers who often need accurate predictions for the effects of climate change in large complex systems. Clearly more work is needed to learn how to accurately predict the ecological responses of species to climate change. Towards that goal, perhaps the best way forward is to conduct more tests like this one.
